### 2023-08-24 sections/1-introduction.tex
some edits to intro

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters. (TODO: Do we need to talk about NR99; it seems like more of the same but 
with removed redundancy; not sure if that is helpful to us...)

@@

In particular, $k$-nearest neighbors (kNN) search is one of the most pervasive supervised classification methods in use~\cite{fix1952discriminatory, cover1967nearest}.
Naive implementations of kNN, whose asymptotic complexity is a function of the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not 
exploit the inherent structure and redundancy often present in big datasets. Approximate search is sufficient for many applications, but the 
need for efficient, \emph{exact} search remains.

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{bingham2001random}, 
clever indexing techniques such as the FM Index~\cite{ferragina2005indexing}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the inherent structure and redundancy often present in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CLAM (Clustered Learning of Approximate Manifolds), and CAKES (CLAM-Accelerated K-nearest-neighbor 
Entropy-scaling Search). CLAM, like CHESS, is a hierarchical clustering algorithm 
that can facilitate a variety of algorithms for big data. Using the tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
$k$-nearest neighbors search in addition to $\rho$-nearest neighbors search.

>>>>

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters.

% (TODO: Do we need to talk about NR99; it seems like more of the same but 
% with removed redundancy; not sure if that is helpful to us...) 
% No we don't... I'll explain why if you want.

@@

In particular, $k$-nearest neighbors (kNN) search is one of the most pervasive supervised classification methods in use~\cite{fix1952discriminatory, cover1967nearest}; most typically classification relies upon a majority vote among the $k$ nearest neighbors to a query.
Naive implementations of kNN, whose time complexity is linear in the dataset's cardinality, 
prove prohibitively slow for large datasets. While algorithms for fast kNN on large datasets exist, most of them are approximate and do not 
exploit the inherent structure and redundancy often present in big datasets. Approximate search is sufficient for many applications, but the 
need for efficient, \emph{exact} search remains.
In particular, while approximate kNN may agree with exact kNN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$.

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{bingham2001random}, 
clever indexing techniques such as the FM Index~\cite{ferragina2005indexing}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the inherent structure and redundancy often present in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated K-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered} is a refinement of the CHESS clustering algorithm. Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
$k$-nearest neighbors search in addition to $\rho$-nearest neighbors search.
